Canonical rings of rational divisors on projective space and Hirzebruch surfaces

### 2015-07-16 surfaces-write-up/surface-write-up.tex
edited projective preliminaries

\begin{prop}
\label{prop:cone-generation}
Let $n \in \bz,$ let $\alpha_1, \ldots, \alpha_n \in \bq$, and let $a_i, b_i \in \bz$ with $1 \leq i \leq n.$ Define
\begin{align*}
	\Sigma = \{(d, c_1, \ldots, c_n) \in \bz^{n+1} : c_i \geq - d \alpha_i,1 \leq i \leq n \text{ and } \sum_{i=1}^{n}a_i = \sum_{i=1}^{n}b_i = 0\}
\end{align*}

\noindent
Suppose $e_1, \ldots, e_t \in \Sigma$ with $e_i = (\pdeg_i, c_1^i,
\ldots, c_n^i)$ are a set of {\bf extremal rays} of $\Sigma$,
meaning that the $\Sigma$ is contained in the $R_{\geq 0}$ span of
$e_1, \ldots, e_t$.  Then, as a semigroup, $\Sigma$ is generated by
elements whose first coordinate is less than $\sum_{i = 1}^{t} \pdeg_i$
. Furthermore, every element $\sigma \in \Sigma$ can be written in
a canonical form $\lambda + \sum_{i=1}^{t} w_i e_i,$ with the first
coordinate of $\lambda$ less than $\sum_{i=1}^{t}\pdeg_i$ and $w_1,
\ldots, w_t \in \bz_{\geq 0}$.
\end{prop}

@@

\begin{lem}
\label{lem:composite-map}
Let $D = \sum_{i=1}^{n}\alpha_i D_i$ where $D_i = V(f_i)$. Suppose we have a surjection $\phi: \bk[x_1,x_2,\ldots, x_r] \rightarrow R_D$ given by $x_i \mapsto p_i(f_1, \ldots, f_n),$ where $p_i$ is a monomial in $f_1,\ldots, f_n$. Let $\bida_1, \ldots, \bida_n, \bidb_1, \ldots, \bidb_n \in \bz_{\geq 0}.$ Then, define
\begin{align*}
	\Sigma = \langle u^d z_1^{c_1} \cdots z_n^{c_n} : c_i \geq -d \alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{n} \bidb_i c_i. \rangle 
\end{align*}
In this case, we can factor $\phi$ through maps $\chi, \psi$ defined by
\todo{Make this have mapsto on bottom}
\[
\begin{tikzcd}
\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
x_i \ar r & u^{d_i}z_1^{c_{i1}} \cdots z_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}
\end{tikzcd}
\]
Assuming $\chi$ is surjective, the minimal degree of generation of $\ker \phi$ is at most the maximum of the minimal degree of generation of $\ker \chi$ and the minimal degree of generation of $\ker \psi$.
\end{lem}
\begin{proof}
Note that we have an exact sequence
\[
\begin{tikzcd}
0 \ar r & \ker \chi \ar r & \ker \phi \ar r & \ker \psi \ar r & 0.
\end{tikzcd}
\]
To see this, note that the left map is an inclusion because $\phi = \chi \circ \psi$. Next, the right map, given by $\chi|_{\ker \phi}:\ker \phi \rightarrow \ker \psi$ is a surjection, again because $\phi = \chi \circ \psi$. The composition of the two maps is $0$ by definition of the kernel of $\chi$. Finally, if $a \in \ker \phi$ and $a$ lies in the kernel of the right map, then by definition $\chi(a) = 0$.
\end{proof}

\begin{lem}
\label{lem:bound-ker-chi}
Retaining the notation of Lemma ~\ref{lem:composite-map}, if $\Sigma$ has extremal rays $e_1,\ldots, e_r$ in degrees $d_1, \ldots, d_r$ then $\ker \chi$ has a system of generators in degrees at most $2(\sum_{i=1}^{r}d_r-1)$.
\end{lem}
\begin{proof}
Since $e_1, \ldots, e_r$ are extremal rays, by Proposition ~\ref{prop:cone-generation}, every element $\sigma \in \Sigma$ can be written in a canonical form $\lambda + \sum_{i=1}^{r}e_i$. Now, letting $\lambda_0 := 0,\lambda_1, \ldots, \lambda_m$ be all elements of $\Sigma$ so that we can write $\lambda_i = \sum_{i=1}^{r}\bida_i e_i$ with $0 \leq \bida_i < 1$ for all $i$. Then, for any $1 \leq i \leq j \leq m,$ we can write $\lambda_i + \lambda_j$ in the above canonical form, yielding a relation in degree at most $\deg \lambda_i + \deg \lambda_j \leq 2 \cdot \left( \sum_{i=1}^{r}d_r -1 \right).$ Furthermore, these relations generate all relations, as one can apply a sequence of these relations to put any $\sigma \in \Sigma$ into canonical form.
\end{proof}

@@

Write
\begin{align*}
	D = \sum_{i=0}^{n}\alpha_i D_i.
\end{align*}
where $\alpha_i \in \bq$ and $\deg D_i = \bida_i$. Let $f_i$ be Cartier divisors such that $D_i = V(f_i)$. \todo{Should this say $D_i$ is a Cartier divisor.  Also should we say $\deg(f_i)= \bida_i$ rather than $\deg(D_i)$.}
We shall further assume that (after reordering) $f_0,\ldots, f_{m}$ form a basis for degree one rational functions on $\bp^m$. \todo{Peter: Make reference to ghost points here?}

>>>>

\begin{prop}
\label{prop:cone-generation}
Let $n \in \bz,$ let $\alpha_1, \ldots, \alpha_n \in \bq$, and let $a_i, b_i \in \bz$ with $1 \leq i \leq n.$ Define
\begin{align*}
	\Sigma = \left \{(d, c_1, \ldots, c_n) \in \bz^{n+1} : c_i \geq - d \alpha_i,1 \leq i \leq n \text{ and } \sum_{i=1}^{n}a_i = \sum_{i=1}^{n}b_i = 0 \right \}
\end{align*}

\noindent
Suppose $e_1, \ldots, e_t \in \Sigma$ with $e_i = (\pdeg_i, c_1^i,
\ldots, c_n^i)$ are a set of {\bf extremal rays} of $\Sigma$,
meaning that the $\Sigma$ is contained in the $R_{\geq 0}$ span of
$e_1, \ldots, e_t$.  Then, as a semigroup, $\Sigma$ is generated by
elements whose first coordinate is less than $\sum_{i = 1}^{t} \pdeg_i$
. Furthermore, every element $\sigma \in \Sigma$ can be written in
a canonical form $\lambda + \sum_{i=1}^{t} w_i e_i,$ with $\lambda = \sum_{i=1}^{r}s_i e_i$ with $0 \leq s_i < 1,$ the first
coordinate of $\lambda$ less than $\sum_{i=1}^{t}\pdeg_i,$ and $w_1,
\ldots, w_t \in \bz_{\geq 0}$.
\end{prop}

@@

\begin{lem}
\label{lem:composite-map}
Let $D = \sum_{i=1}^{n}\alpha_i D_i$ where $D_i = V(f_i)$. Suppose we have a surjection $\phi: \bk[x_1,x_2,\ldots, x_r] \rightarrow R_D$ given by $x_i \mapsto p_i(f_1, \ldots, f_n),$ where $p_i$ is a monomial in $f_1,\ldots, f_n$. Let $\bida_1, \ldots, \bida_n, \bidb_1, \ldots, \bidb_n \in \bz_{\geq 0}.$ Then, define
\begin{align*}
	\Sigma = \langle u^d z_1^{c_1} \cdots z_n^{c_n} : c_i \geq -d \alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{n} \bidb_i c_i = 0 \rangle. 
\end{align*}
In this case, we can factor $\phi$ as a composition of $\chi$ and $\psi$ defined by
\todo{Make this have mapsto on bottom}
\[
\begin{tikzcd}
\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
x_i \ar r & u^{d_i}z_1^{c_{i1}} \cdots z_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]
Assuming $\chi$ is surjective, the minimal degree of generation of $\ker \phi$ is at most the maximum of the minimal degree of generation of $\ker \chi$ and the minimal degree of generation of $\ker \psi$.
\end{lem}
\begin{proof}
First, we show there is an exact sequence
\[
\begin{tikzcd}
0 \ar {r} & \ker \chi \ar{r}{\rho} & \ker \phi \ar {r}{\tau} & \ker \psi \ar r & 0.
\end{tikzcd}
\]
To see this, note that $\rho$ is injective and $\tau$ is surjective\ because $\phi = \chi \circ \psi$. Additionally, $\tau \circ \rho = 0$ by definition of the kernel of $\chi$. Finally, if $a \in\ker \tau$, then by definition $\chi(a) = 0,$ so $a \in \im \rho$.

This shows that lifts of generators of $\ker \psi$ together with images of generators of $\ker \chi$ generate all of $\ker \phi,$ as desired. 
\end{proof}

\begin{lem}
\label{lem:bound-ker-chi}
Retaining the notation of Lemma ~\ref{lem:composite-map}, if $\Sigma$ has extremal rays $e_1,\ldots, e_r$ in degrees $d_1, \ldots, d_r$ then $\ker \chi$ is generated in degrees up to $2(\sum_{i=1}^{r}d_r-1)$.
\end{lem}
\begin{proof}
Since $e_1, \ldots, e_r$ are extremal rays, by Proposition ~\ref{prop:cone-generation}, every element $\sigma \in \Sigma$ can be written in a canonical form $\lambda + \sum_{i=1}^{r}e_i$. Now, let $\lambda_0 := 0,\lambda_1, \ldots, \lambda_m$ be all elements of $\Sigma$ which can be written in the form $\lambda_i = \sum_{i=1}^{r}r_i e_i$ with $0 \leq r_i < 1.$ Then, for any $1 \leq i \leq j \leq m,$ we can write $\lambda_i + \lambda_j$ in the above canonical form, yielding a relation in degree at most $\deg \lambda_i + \deg \lambda_j \leq 2 \cdot \left( \sum_{i=1}^{r}d_r -1 \right).$ Furthermore, these relations generate all relations, as one can apply a sequence of these relations to put any $\sigma \in \Sigma$ into canonical form $\sigma = \lambda + \sum_{i=1}^{r}w_i e_i$ as in Proposition ~\ref{prop:cone-generation}.
\end{proof}

@@

Write
\begin{align*}
	D = \sum_{i=0}^{n}\alpha_i D_i.
\end{align*}
where $n \in \bz$, $\alpha_i \in \bq$, and $\deg D_i = \bida_i$. Let $D_i$ be Cartier divisors such that $D_i = V(f_i)$.
We shall further assume that (after reordering) $f_0,\ldots, f_{m}$ form a basis for degree one rational functions on $\bp^m$.
